---
title: "Transmission Type and Fuel Efficiency"
author: "Alyssa Goldberg"
date: "January 10, 2016"
output:
  html_document: default
  pdf_document:
    fig_caption: yes
    highlight: pygments
  word_document: default
---

#Executive Summary
##Hypotheses and Questions
Using the mtcars dataset from a 1974 Motor Trend US magazine, we examine the following hypotheses:
$H_0$ = There is no significant difference in fuel efficiency between cars with **automatic** and **manual** transmission.
$H_01$ = There are no confounding variables that contribute to the acceptance or rejection of $H_0$. 

1. Is an automatic or manual transmission better for MPG ?
2. If so, quantify the MPG difference between automatic and manual transmissions 

##Summary of Findings  
*All figures are in the appendix*
1. Cars with manual transmission get **7.24** more mpg than automatic transmission when only transmission type is considered
2. Exploring regression models using multivariate analysis strongly suggests that both weight `wt`, number of cylinders `cyl`  and `hp` have a significant confounding effect on fuel efficiency as it relates to transmission type.  
With other variables factored in, cars with manual transmission get an additional **1.8** mpg better than those with automatic transmission.

#Data 
The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models). 

Note: The 1/4 mile time `qsec` is a measure of performance rather than a manufacturing variable and thus we remove qsec from the dataset for the purposes of analysis. 

##Factors and Measurements
`mpg` Miles/(US) gallon  (the measure of,  performance we are interested in), `disp` Displacement (cu.in.), `hp` Gross horsepower, `drat` Rear axle ratio, `wt` Weight (lb/1000), `qsec` 1/4 mile time, `cyl` Number of cylinders, `vs` V/S, `am` Transmission (0 = automatic, 1 = manual), `gear` Number of forward gears, `carb` Number of carburetors.  
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(stats)

library(ggplot2)
library(reshape2)
library(dplyr)
library(car)

library(corrplot)
library(graphics)


data(mtcars)

mtcars$cyl <- factor(mtcars$cyl)
mtcars$vs <- factor(mtcars$vs)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
mtcars$am <- factor(mtcars$am,labels=c('Automatic','Manual'))

data<-mtcars[,-7]
```
```{r echo=FALSE}

fit<-lm(mpg~am, mtcars[,-c(7)])
```
Figure 1 and Table 1 (below) include only `mpg` and `am` indicate that manual transmission has a mean mpg `r round(summary(fit)$coef[2], 3) ` automatic. 

A t-test of `mpg` and `am` can reveal both the confidence interval and p-values for the two factors in the `am` variable:

```{r}
t.test(mtcars[,-7]$mpg~mtcars[,-7]$am)
```

The CI, `r t.test(mpg~am, data, alternative = "two.sided")$conf `, does not contain zero *and* a p-value of `r t.test(mpg~am, data, alternative = "two.sided")$p.value`, significantly better than our 95% confidence level $alpha = 0.05%.$. If the only two variable measured were fuel efficiency and transmission type, we could safely reject $H_0$.

There are 8 additional manufacturing variables in the data set and the adjusted $R^2$ for `mpg~am` is low, only `r round(summary(fit)$adj.r.squared, 3)`, so transmission alone accounts for only `r round(summary(fit)$adj.r.squared*100, 2)`% of the total effect. We need to find a better model.
```{r echo=FALSE}
xt<-summary(fit)$coefficients
```
```{r}
kable(xt, caption = "Table 1: Linear Regression Coefficients for Automatic vs. Manual Transmission")
```

Figure 2 shows that `cyl`, `disp`, `wt` and `hp` all have a negative effect on fuel efficiency. 

#Find the Best Model
##Step Analysis for AIC
We employ R's step() function to find the model with the lowest AIC. (results hidden to limit length of report) and produce a table comparing our three models, `mpg ~ am`, `mpg ~ .` and the result of the step analysis, `mpg ~ cyl + hp + wt + am`.
```{r echo=FALSE, results='hide'}

bestfit<-step(lm(mpg~., mtcars[,-c(7)]), direction = "both")
bestmodel<-bestfit$call
bfform<-bestmodel$formula
bfform<-paste(bfform[c(2,1,3)], collapse = "")
allAIC=extractAIC(lm(mpg ~ ., mtcars[,-7]))
fitAIC=extractAIC(lm(mpg ~ am, mtcars[,-7]))
bestAIC=extractAIC(bestfit)
tblAIC<-as.data.frame(rbind("mpg ~ am"= fitAIC[2],"mpg ~ ." = allAIC[2],"mpg ~ cyl + hp + wt + am" = bestAIC[2]))
names(tblAIC)<-"AIC"
```
```{r}
fitAIC=extractAIC(lm(mpg ~ am, mtcars[,-7]))
allAIC=extractAIC(lm(mpg ~ ., mtcars[,-7]))
bestAIC=extractAIC(lm(mpg ~ cyl + hp + wt + am, mtcars[,-c(7)]))
tblAIC<-as.data.frame(rbind("mpg ~ am"= fitAIC[2],"mpg ~ ." = allAIC[2],"mpg ~ cyl + hp + wt + am" = bestAIC[2]))
names(tblAIC)<-"AIC"
kable(tblAIC, caption = "Comparison of AIC Values of Three Models")
```

The model which includes: `r bfform` has the lowest AIC value of  `r extractAIC(bestfit)[[2]]`  and is recommended as the best fit by the step analysis.  A summary of the best fit model coefficients confirm we are moving in the right direction. The net influence of the `am` variable when `am` = 1 (manual transmission) is `r round(summary(bestfit)$coefficients[6,1], digits = 3)`. We will continue to test this model against the others.
```{r echo=FALSE}
# bestmodel$formula
```
```{r echo=FALSE, results='asis'}
kable(summary(bestfit)$coefficients)
```

##Analysis of Variance
An analysis of variance shows us that number of `cyl`, `wt` and `disp` are significant at a 95% confidence level, while the p-value of `hp` is higher than .05 indicating we *might* reject it as a significant factor. 

```{r}
kable(as.data.frame(summary(aov(mpg~., data))[[1]]))
```

The next step in finding out how closely correlated these confounding variables are. A correlation matrix (figure X) shows that `disp` and `cyl` have a correlation of 90% while `wt` and `disp` are 89%. Both show significantly lower p-values so it is possible that including `disp` to a model might result in "overfit."

```{r echo=FALSE}
mtcars<-mtcars[,-c(7)]
mtcars$cyl<-as.numeric(mtcars$cyl)
c1 <- cor(mtcars$wt, mtcars$cyl)
c2<-cor(mtcars$wt, mtcars$disp)
c3<-cor(mtcars$cyl, mtcars$disp)
c4<-as.data.frame(rbind("wt & cyl" = c1, "wt & disp" = c2, "cyl & disp" = c3))
names(c4)<-"correlation"
```
```{r, results='asis'}
kable(c4, caption = "Correlation among variables with lowest p-value in analysis of variance model")
```
```{r echo=FALSE}
fitall<-lm(mpg~., mtcars[,-c(7)])
fitlesshp <- lm(mpg ~ am + cyl + wt, data = mtcars[,-c(7)])
fitsummary<-data.frame(rbind("mpg ~ ."=summary(fitall)$adj.r.squared, "mpg ~ am"=summary(fit)$adj.r.squared, "mpg ~ cyl  + wt + am" = summary(fitlesshp)$adj.r.squared, "mpg ~ cyl + hp + wt + am " = summary(bestfit)$adj.r.squared))
names(fitsummary)<-"Adj. R-squared"
```
We compare several models to determine which one is truly "best" and the fourth model which retains `hp` does have the highest $R^2$: `r summary(bestfit)$adj.r.squared`, accounting for `r round((summary(bestfit)$adj.r.squared)*100)`% of the variability in fuel efficiency.

```{r}
kable(fitsummary)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
firstfit<-lm(mpg~am, mtcars[,-c(7)])
fithp<-lm(mpg~am + hp, mtcars[,-7])
final<-anova(firstfit, bestfit)
final$PRF<-final$`Pr(>F)`
```


An ANOVA test of the best fit model and the original model provides a p-value of `r round(final$PRF[2], 10)` indicating that we can reject $H_01$-that number of cylinders, weight and horsepower are not confounding factors (more cylinders = lower `mpg`, higher weight = lower mpg, higher horsepower=lower mpg ) to transmission type as regards fuel efficiency.

When number of cylinders, weight and horsepower are included as confounding variables, the net increase in fuel efficiency for an automobile with automatic transmission is ** `r round(summary(bestfit)$coefficients[6,1], digits = 3)`**

The table below generated with R's bootstrap function lists the relative importance of each variable and how much it contributes to the final $R^2$ value of 84%.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(relaimpo)

bs <- boot.relimp(bestfit, b = 10, type = "lmg", rank = TRUE,diff = TRUE, rela = TRUE)
# bs2=booteval.relimp(bs)
bs2=booteval.relimp(bs, bty="perc", level = 0.95, sort = TRUE, typesel = "lmg")
btbl<-as.data.frame(as.numeric(bs2@mark[1:4])*100)
names(btbl)<-c("% Contribution")
row.names(btbl)<-c("cyl", "hp", "wt", "am")
```

```{r}
kable(btbl,caption = "% contribution to R-squared value of best fit model" )
```

##Residuals  
The Residual Plots (Figure 4) confirm our model in the following ways:
1. Residuals vs. Fitted plot points are randomly distributed confirming independence.
2. Points on the Normal Q-Q plot hug the normal line with some outliers in the tails confirming normal
distribution of the residuals.
3. The Scale-Location plot displays a regular band pattern of points confirming constant variance.

#Appendix
figure 1: Boxplot mpg ~ am (mpg vs transmission type)  
```{r boxplot, echo=FALSE}
ggplot(mtcars, aes(y=mpg, x=am, col=am), cex=.5) +
        geom_boxplot() +
        geom_jitter() +
        labs(title="Fuel Efficiency (MPG) vs Transmission Type", x="", y="MPG") +
        theme(legend.position="none")
```

Figure 2: Pairs Plot 
```{r pairs, echo=FALSE}

pairs(x = mtcars, lower.panel = panel.smooth, upper.panel = NULL)
```

Figure 3: Correlation Matrix  
```{r corrplot, echo=FALSE}
par(mfrow=c(1,1))
data(mtcars)
corrplot(cor(mtcars), method="number", type="lower", addCoefasPercent = TRUE)
```

Figure 4: Residual Plots  
```{r echo=FALSE}
par(mfrow=c(1,1))
par(mfrow=c(2,2))
plot(bestfit)
```

